---
title: "intestine_import"
format: html
---

## Import data

Through the terminal, the data has been downloaded using curl (on the server), where its been extracted into sample specific directories. The files in each directory is then used to create Seurat objects per sample (with RNA and ATAC assays). These Seurat objects are written to "/home/projects/exam_2026_22102/group2/intestine_data/sample_seurats"

```{r}
suppressPackageStartupMessages({
    library(Seurat)
    library(Signac)
    library(GenomeInfoDb)
    library(stringr)
    library(rtracklayer)
    library(GenomicRanges)
  })

source("./functions.R")
```

```{r}
B006_seurat <-rna_atac_integration_signac("B006")
```

```{r}
saveRDS(B006_seurat, file="./B006_seurat.rds")
```

## Running all and saving rds.

```{r}
save_path <- "/home/projects/exam_2026_22102/group2/intestine_data/sample_seurats/" #dir to save individual sample seurat objects.
prefixes <- c("B006", "B008", "B010", "B011", "B012")

for (prefix in prefixes) {
  seurat_obj <- rna_atac_integration_signac(prefix)
  
  # Save Seurat object as RDS file, using the prefix in the filename
  saveRDS(seurat_obj, file = str_c(save_path, prefix, "_seurat.rds"))
  
  print(str_c("Completed !", prefix))
}
```

## Manual loading

```{r}
############################################
## RNA + ATAC integration with Signac
############################################

############################################
## Paths and project info
############################################
proj_prefix <- "B006"
message(str_c("Starting ",proj_prefix))

intestine_path <- "/home/projects/exam_2026_22102/group2/intestine_data/"

#Regex patterns
matrix_pattern <- ".*_matrix\\.mtx\\.gz$"
features_pattern <- ".*_features\\.tsv\\.gz$"
barcodes_pattern <- ".*_barcodes\\.tsv\\.gz$"
fragments_pattern <- ".*_atac_fragments\\.tsv\\.gz\\.tbi$"
#peaks_pattern <- "stromal_peaks\\.bed\\.gz$"

rna_matrix   <- return_regex_filepath(intestine_path, proj_prefix, matrix_pattern)
rna_features <- return_regex_filepath(intestine_path, proj_prefix, features_pattern)
rna_barcodes <- return_regex_filepath(intestine_path, proj_prefix, barcodes_pattern)
fragments_file <- return_regex_filepath(intestine_path, proj_prefix, fragments_pattern)

peaks_file <- str_c(intestine_path, "stromal_peaks.bed.gz") #only in one place

# Check if files are recognizes
#message("rna_matrix", rna_matrix)
#message("rna_features", rna_features)
#message("rna_barcodes", rna_barcodes)
message("fragments_file", fragments_file)
#message("peaks_file", peaks_file)

############################################
## 1. Load RNA counts and create Seurat object
############################################
message("Loading RNA counts and creating Seurat object")
counts <- ReadMtx(
  mtx = rna_matrix,
  features = rna_features,
  cells = rna_barcodes
)

seurat_obj <- CreateSeuratObject(
  counts = counts,
  project = proj_prefix
)
```

```{r}
# Extract the gene names
gene_names <- counts@Dimnames[[1]]

# Find which names are duplicates
dup_logic <- duplicated(gene_names)
duplicate_genes <- gene_names[dup_logic]

# Count how many duplicates there are
length(duplicate_genes)

# See the actual names of the duplicates
unique(duplicate_genes)
```

```{r}
############################################
## 2. Load ATAC peaks (space-separated file)
############################################
message("Loading ATAC peaks")
peaks_df <- read.table(
  peaks_file,
  header = FALSE,
  sep = "",
  stringsAsFactors = FALSE
)

peaks_df <- peaks_df[, 1:3]
colnames(peaks_df) <- c("chr", "start", "end")

peaks <- makeGRangesFromDataFrame(
  peaks_df,
  seqnames.field = "chr",
  start.field = "start",
  end.field = "end",
  keep.extra.columns = FALSE
)

peaks <- keepStandardChromosomes(peaks, pruning.mode = "coarse")
peaks <- peaks[width(peaks) > 20]

```

Investigating fragment error

```{r}
fragments_pattern <- ".*_atac_fragments\\.tsv\\.gz$"
fragments_file <- return_regex_filepath(intestine_path, proj_prefix, fragments_pattern)
file <- gzfile(fragments_file, "rt")
```

```{r}
# Check Seurat object barcodes
head(colnames(seurat_obj))

# Check Fragment file barcodes (using tabix via Signac/R)
# Note: You can also check this in terminal using: zcat fragments.tsv.gz | head -n 20
#head(read.table(file, nrow = 10)$V4)

head(read.table(file, nrow = 10))
```

```{r}
############################################
## 3. Create Fragment object
############################################
message("Creating Fragment object")
fragments <- CreateFragmentObject(
  path = fragments_file,
  cells = colnames(seurat_obj)
)
```

```{r}
############################################
## 4. Create ATAC count matrix
############################################
message("Creating ATAC count matrix")
atac_counts <- FeatureMatrix(
  fragments = fragments,
  features = peaks,
  cells = colnames(seurat_obj)
)

```

```{r}
############################################
## 5. Create ChromatinAssay
############################################
message("Creating ChromatinAssay")
chrom_assay <- CreateChromatinAssay(
  counts = atac_counts,
  fragments = fragments,
  genome = "hg38"  # change to mm10 if mouse
)
```

```{r}
############################################
## 6. Add ATAC assay to Seurat object
############################################
message("Adding ATAC assay to Seurat object")
seurat_obj[["ATAC"]] <- chrom_assay
DefaultAssay(seurat_obj) <- "RNA"

message(str_c("Completed ",proj_prefix))
seurat_obj
```

```{r}
save_path <- "/home/projects/exam_2026_22102/group2/intestine_data/sample_seurats/"
saveRDS(seurat_obj, file = str_c(save_path, proj_prefix, "_seurat.rds"))
```
