---
title: "02_data_processing"
format: html
---

```{r}
suppressPackageStartupMessages({
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(harmony)
library(tidyr)
#library(multtest)
#library(metap)
library(tibble)
library(Matrix)
library(scDblFinder)
library(scater)
})

source("./paths.R")
#source("./functions.R")
```

## Load and combine sample data

```{r}
#reading the seruat objects
# B006 <- readRDS(str_c(sample_seurat_path, "B006_seurat.rds"))
# B008 <- readRDS(str_c(sample_seurat_path, "B008_seurat.rds"))
# B010 <- readRDS(str_c(sample_seurat_path, "B010_seurat.rds"))
# B011 <- readRDS(str_c(sample_seurat_path, "B011_seurat.rds"))
# B012 <- readRDS(str_c(sample_seurat_path, "B012_seurat.rds"))


# Adding meta data of each donor to thei seurat object
# B006$donor <- "B006"
# B008$donor <- "B008"
# B010$donor <- "B010"
# B011$donor <- "B011"
# B012$donor <- "B012"


# EC <- merge(
#   x = B006,
#   y = list(B008, B010, B011, B012),
#   add.cell.ids = c("B006","B008","B010","B011", "B012"),
#   project = "AllDonors"
# )
# 
# 
# # checking if meta data was perserved
# table(EC$donor)
# 
# # Joing layers so count matrix and such is pooled
# EC <- JoinLayers(EC)
# 
# # Saving the seurat 
# saveRDS(EC, file = paste0(EC_seurats_path, "EC_merged_seurat.rds"))
# 
# dim(EC) # dim before subsetting, 545083 3569104

```

## Loading merged seurat

```{r}
EC <- readRDS(paste0(EC_seurats_path, "EC_merged_seurat.rds"))
```

Removing spatial information:

```{r}
# saving counts
counts <- GetAssayData(EC, layer= 'counts')

#Saving spatial 
spatial_genes <-  rownames(EC)[grepl(":", rownames(EC))]

#adding spatial genes:
#EC <- EC[spatial_genes, ] 

#Remving spatial information
keep_genes <- rownames(EC)[!grepl(":", rownames(EC))]

# Subset the Seurat object
EC[["RNA"]] <- subset(EC[["RNA"]], features = keep_genes)

#EC <- subset(EC, features = keep_genes) #asbj: subset function was more safe than using []
```

# QC of the data

## Percentages of mito and ribosomal

```{r}
#Mitochondrial genes
EC@meta.data[["percent.mt"]] <- PercentageFeatureSet(EC, pattern = "^MT-")

#Ribosomal genes proportions
EC[["percent.ribo"]] <- PercentageFeatureSet(EC, pattern = "^RPS|^RPL|^rps|^rpl")

```

Converting NA to 0:

```{r}
features <- c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo")

for (f in features) {
  if (f %in% colnames(EC@meta.data)) {          # check it exists
    x <- EC@meta.data[, f]                       # extract as vector
    x[is.na(x) | is.infinite(x)] <- 0            # replace NA/Inf with 0
    EC@meta.data[, f] <- x                        # write back
  } else {
    warning(paste("Feature not found in metadata:", f))
  }
}
```

Checking for no Na values for QC metric, should contain only zeros

```{r}
colSums(is.na(EC@meta.data[, c(
  "nFeature_RNA",
  "nCount_RNA",
  "percent.mt",
  "percent.ribo"
)]))
```

Making QC metric ViolinPlot

```{r}

# making identity same for all
Idents(EC) <- factor(rep("all-cells", ncol(EC)))
levels(Idents(EC))

# Restoring the identity: 
# Idents(EC) <- EC$donor

# Visualize QC metrics as a violin plot
p <- VlnPlot(EC, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo"), ncol = 4, group.by = NULL)
p
```

Proposed QC cutoffs:

-   nFeature_RNA 200\< x \<35,000

-   percent.mt: 8% = 0.08, as we do expect a bit higher metabolic activity (see litt. reference in pp)

-   percent.ribo: 50%

### Implementing QC Filtering

```{r}
cells_before <- ncol(EC)
counts_before <- as.data.frame(table(EC$orig.ident))
dist_sample_table_before <- table(EC$orig.ident)

cells_to_keep <- WhichCells(EC, expression = 
                              nFeature_RNA > 200 &
                              nFeature_RNA < 35000 &
                              percent.mt < 8 &
                              percent.ribo < 50)

# intersect with ATAC cells
cells_to_keep <- intersect(cells_to_keep, colnames(EC[["ATAC"]]))

# subset
EC <- subset(EC, cells = cells_to_keep)

# EC <- subset(
#   x = EC,
#   subset = nFeature_RNA > 200 &
#            nFeature_RNA < 35000 &
#            percent.mt < 8 &
#            percent.ribo < 50)

cells_after <- ncol(EC)
cells_lost <- cells_before - cells_after
counts_after <- as.data.frame(table(EC$orig.ident))
dist_sample_table_after <- table(EC$orig.ident)

print(paste("Cells lost:", cells_lost))
print(paste("Percentage lost:", round((cells_lost/cells_before)*100, 2), "%"))
```

Distribution of cells lost per sample

```{r}
print(dist_sample_table_before)
print(dist_sample_table_after)
```

```{r}
# 2. Merge them together
plot_data <- merge(counts_before, counts_after, by = "Var1")
colnames(plot_data) <- c("Sample", "Before", "After")

# 3. Reshape for ggplot
plot_data_long <- pivot_longer(plot_data, cols = c("Before", "After"), 
                               names_to = "Status", values_to = "CellCount")

# 4. Create the plot
ggplot(plot_data_long, aes(x = Sample, y = CellCount, fill = Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Cell Counts Before vs. After Filtering",
       y = "Number of Cells",
       x = "Sample ID") +
  scale_fill_manual(values = c("Before" = "grey80", "After" = "steelblue"))
```

```{r}
# Calculate percentage kept
plot_data$PercentKept <- (plot_data$After / plot_data$Before) * 100

ggplot(plot_data, aes(x = Sample, y = PercentKept, fill = Sample)) +
  geom_bar(stat = "identity") +
  # Add the labels on top of the bars
  geom_text(aes(label = paste0(round(PercentKept, 1), "%")), 
            vjust = -0.5,             # Move text slightly above the top of the bar
            size = 3) +               # Adjust font size
  theme_minimal() +
  labs(title = "Retention Rate per Sample",
       y = "% of Cells Remaining",
       x = "Sample ID") +
  ylim(0, 110) +                      # Increase y-limit slightly to give space for labels
  scale_fill_brewer(palette = "Set2") # Optional: nicer color palette
```

# Random subsetting data

```{r}
cells.use <- unlist(
  lapply(
    split(intersect(Cells(EC), colnames(EC[["ATAC"]])), EC$donor),
    function(x) sample(x, 8000)
  )
)
EC <- subset(EC, cells = cells.use)

```

```{r}

set.seed(123) # seed for reproducibility

cells.use <- unlist(
  lapply(
    split(Cells(EC), EC$donor),
    function(x) sample(x, 8000)
  )
)

EC <- subset(EC, cells = cells.use)

dim(EC) # dim subsetting
```

Recapping the subsetting of data:

```{r}
cells_final <- ncol(EC)
counts_final <- as.data.frame(table(EC$orig.ident))
cells_lost_final <- cells_before - cells_final

print(paste("Cells lost (Final):", cells_lost_final))
print(paste("Final percentage lost:", round((cells_lost_final/cells_before)*100, 2), "%"))
```

Base count lost plot

```{r}
# 2. Merge them together
plot_data <- merge(counts_before, counts_final, by = "Var1")
colnames(plot_data) <- c("Sample", "Before", "After")

# 3. Reshape for ggplot
plot_data_long <- pivot_longer(plot_data, cols = c("Before", "After"), 
                               names_to = "Status", values_to = "CellCount")

# 4. Create the plot
ggplot(plot_data_long, aes(x = Sample, y = CellCount, fill = Status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Cell Counts Before vs. After Filtering",
       y = "Number of Cells",
       x = "Sample ID") +
  scale_fill_manual(values = c("Before" = "grey80", "After" = "steelblue"))
```

Percentage lost plot

```{r}
# Calculate percentage kept
plot_data$PercentKept <- (plot_data$After / plot_data$Before) * 100

ggplot(plot_data, aes(x = Sample, y = PercentKept, fill = Sample)) +
  geom_bar(stat = "identity") +
  # Add the labels on top of the bars
  geom_text(aes(label = paste0(round(PercentKept, 1), "%")), 
            vjust = -0.5,             # Move text slightly above the top of the bar
            size = 3) +               # Adjust font size
  theme_minimal() +
  labs(title = "Retention Rate per Sample",
       y = "% of Cells Remaining",
       x = "Sample ID") +
  ylim(0, 110) +                      # Increase y-limit slightly to give space for labels
  scale_fill_brewer(palette = "Set2") # Optional: nicer color palette
```

# Normalization, variables features and scaling

Now we're left with a filtered EC object based on QC cutoffs then subsetting for computational feasibility

```{r}
#| warning: false
#| message: false

EC <- NormalizeData(EC)
EC <- FindVariableFeatures(EC)
EC <- ScaleData(EC)
EC <- RunPCA(EC)
```

Detecting viable PCs with Elbowplot

```{r}
ElbowPlot(EC) # 1:9 looks good
```

Running and storing naive UMAP embedding

```{r}
EC <- FindNeighbors(EC, dims = 1:9)
EC <- FindClusters(EC, algorithm = 4, random.seed = 40)
EC <- RunUMAP(EC, dims=1:9)
```

## Doublets investigation

Checking for doublets without filtering

```{r}
# converting to sce element  
sce_EC <- as.SingleCellExperiment(EC)    
# perorming doublet finder  
set.seed(42) 
sce_EC.dbl <- scDblFinder(sce_EC ,samples="donor", clusters=colLabels(sce_EC)) 
```

```{r}
sce_EC.dbl <- logNormCounts(sce_EC.dbl)  
# Check that the assay exists now 
assayNames(sce_EC.dbl)
```

```{r}
# plot UMAP of potential doublets 
sce_EC.dbl <- runUMAP(sce_EC.dbl, exprs_values = "logcounts")  
plotUMAP(sce_EC.dbl, colour_by="scDblFinder.score")
```

```{r}
# Finding seurat class for doublet  
table(sce_EC.dbl$scDblFinder.class)
```

```{r}
plotColData(sce_EC.dbl, y="scDblFinder.score", colour_by=("scDblFinder.class"))
```

# 

# Saving final Seurat object

```{r}
# Saving with a simple name "EC"
saveRDS(EC, file = paste0(EC_seurats_path, "02_EC.rds"))
```
