---
title: "03_sample_integration"
format: html
---

```{r}
# Load libraries
#| echo: false
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(harmony)
  library(tidyr)
  library(multtest)
  library(metap)
  library(tibble)
  library(future)
})

source("./paths.R")
#source("./functions.R")

EC <- readRDS(paste0(EC_seurats_path, "02_EC_v2.rds"))
DefaultAssay(EC) <- "RNA"
```

## Before integration

For sample integration it is our aim to perform horizontal integration across donors. This enables us to compare donors and capture biological variation (instead of technical). We'll take a look at the UMAP reduction to get a grasp of the cluster formation before integration.

```{r}
# Initial look at clustering (before integration)
DimPlot(EC, reduction="umap", group.by = "donor")
```

This plot reveals that....

Before any integration can take place, we split the data sets into five Seurat Objects by using condition and go through pre-processing one more time for each object.

```{r}
# Split the data
EC[["RNA"]] <- split(EC[["RNA"]], f = EC$donor)
```

```{r}
EC <- NormalizeData(EC)
EC <- FindVariableFeatures(EC)
EC <- ScaleData(EC)
EC <- RunPCA(EC)
```

To determine the number PCs to use, we plot and ElbowPlot()

```{r}
ElbowPlot(EC)
```

**Integrate layers:**

```{r}
# Anchor-based CCA integration
# REMEMBER TO DETERMINE PCS
EC <- IntegrateLayers(object = EC, 
                       method = CCAIntegration,
                       orig.reduction = "pca", 
                       new.reduction = "CCA_Integration",
                       dims = 1:15, 
                       verbose = FALSE
)

```

```{r}
options(future.globals.maxSize = 6 * 1024^3)

EC <- IntegrateLayers(object = EC, 
                       method = RPCAIntegration ,
                       orig.reduction = "pca", 
                       new.reduction = "RPCA_Integration",
                       dims = 1:15, 
                       verbose = FALSE
)


```

```{r}

EC <- IntegrateLayers(object = EC, 
                       method = HarmonyIntegration ,
                       orig.reduction = "pca", 
                       new.reduction = "harmony_Integration",
                       dims = 1:15, 
                       verbose = FALSE)

EC[["RNA"]] <- JoinLayers(EC[["RNA"]])
```

After integration, we will run UMAP reduction again and visualize the new clustering.

```{r}
# Run UMAP
EC <- RunUMAP(EC, 
               dims = 1:15, 
               reduction = "CCA_Integration", 
               reduction.name = "umap_cca")


EC <- RunUMAP(EC, 
               dims = 1:15, 
               reduction = "RPCA_Integration", 
               reduction.name = "umap_rpca")


EC <- RunUMAP(EC, 
               dims = 1:15, 
               reduction = "harmony_Integration", 
               reduction.name = "umap_harmony")
```

### Visualize integrated clusters

```{r}
p1 <- DimPlot(EC, reduction = 'umap', group.by = 'donor') + ggtitle("UMAP non-integrated data")
p2 <- DimPlot(EC, reduction = "umap_cca", group.by = "donor") + ggtitle("UMAP CCA")
p3 <- DimPlot(EC, reduction = "umap_rpca", group.by = "donor") + ggtitle("UMAP RPCA")
p4 <- DimPlot(EC, reduction = "umap_harmony", group.by = "donor") + ggtitle("UMAP Harmony")

wrap_plots(p1, p2, p3, p4, nrow = 2) + plot_layout(guides = "collect")
```

# Saving final Seurat object

```{r}
# Saving with a simple name "EC"
# The EC now contain integrated data along with a new reduction; umap_cca
saveRDS(EC, file = paste0(EC_seurats_path, "03_EC.rds"))
```
