---
title: "04_cell_type_annotation"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(multtest)
  library(metap)
  library(tibble)
  library(clustree)
  library(SingleR)
})

source("./paths.R")
#source("./functions.R")

EC <- readRDS(paste0(EC_seurats_path, "03_EC.rds"))
```

#Maunal Cell annotation

### Cell clustering

For cell type annotation, we will perform clustering first to gain insight into the local relationships of cell types.

```{r}
# Identify neighbors based on the CCA integration and perform clusering
EC <- FindNeighbors(EC,
                         dims = 1:15, 
                         reduction = "harmony_Integration", 
                         graph.name = "harmony_snn")

EC <- FindClusters(EC, 
                        resolution = c(0.4,0.6,0.8,1.0,1.2), 
                        algorithm = 4,
                        random.seed = 1,
                        graph.name = "harmony_snn")
```


### Clustering optimization

Clustering optimization is necessary to come to a clustering of cell types that we can confidently continue our analysis on. For optimization, we will test different resolutions (k) and determine the lowest resolution that provides a good clustering. To investigate this, we will create a clustree.

```{r}
clustree(EC, prefix = "harmony_snn_res.")
```

```{r}
# Set Idents to the resolution that was best
Idents(EC) <- "harmony_snn_res.0.8" 
```


#Tile plot

```{r}


df <- data.frame(
  cluster = EC@meta.data$harmony_snn_res.0.8,
  donor = EC@meta.data$donor
)

summary_df <- df %>%
  group_by(cluster, donor) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(fraction = count / sum(count))

```


```{r}
ggplot(summary_df, aes(x = donor, y = cluster, fill = fraction)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(fraction, 2)), color = "black", size = 3) + 
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(x = "Donor", y = "Cluster", fill = "Fraction") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```





### Cell type assignment


```{r}

# Definér hvilke donorer der er Disease
disease_donors <- c("B008", "B011")

# Opret ny metadata-kolonne
EC$Disease_status <- ifelse(
  EC$donor %in% disease_donors,
  "Disease",
  "Healthy"
)

# Tjek resultatet
table(EC$Disease_status, EC$donor)


```


**Manual marker cell type assignment**

It is also possible to utilize `FindConsveredMarkers()` and `FindMarkers()` for identifying genes that stand out in a cluster.

```{r}
# Compare one cluster to all the other clusters
# USE IF A CLUSTER IS STANDING OUT FROM OTHERS
clusterX_conserved_markers <- FindConservedMarkers(EC,
                                                   ident.1 = 10,
                                                   grouping.var = "Disease_status",
                                                   only.pos = TRUE, min.pct = 0.25,
                                                   min.diff.pct = 0.25, 
                                                   logfc.threshold = 0.25)
```

```{r}
# Making feature grid plot for each of the features that are found

head(clusterX_conserved_markers)

head(rownames(clusterX_conserved_markers))

```


#Annotate the cell types from the found features

```{r}
# 1. Create a mapping dictionary named "cell_typ_dict" like below.
cell_typ_dict <- c( "1" = "CELL_TYPE_A", # replace these with real cell types.
                     "2" = "CELL_TYPE_B",
                     "3" = "CELL_TYPE_C",
                     "4" = "CELL_TYPE_A",
                     "5" = "CELL_TYPE_A",
                     "6" = "CELL_TYPE_B",
                     "7" = "CELL_TYPE_B",
                     "8" = "CELL_TYPE_C",
                     "9" = "CELL_TYPE_C",
                     "10" = "CELL_TYPE_D",
                     "11" = "CELL_TYPE_D",
                     "12" = "CELL_TYPE_E",
                     "13" = "CELL_TYPE_F",
                     "14" = "CELL_TYPE_F",
                     "15" = "CELL_TYPE_G")
# 2. Map the clusters to the identified cell types and store them as a new column named "cell_type_manual".
pbmc.MTB.cca$cell_type_manual <- unname(cell_typ_dict[as.character(Idents(pbmc.MTB.cca))])

```



**Automatic cell type assignment**

following this tutorial: https://biostatsquid.com/singler-tutorial/

Initially, performing automatic cell type assignment due to not only the speed. but also now the quality of the reference database.

To run automatic cell type assignment the Seurat object has to be a SingleCellExperiment (SCE) object

First of all we need a cell annotation reference dataset

```{r}
raw_counts <- LayerData(seu, assay = "RNA", layer = 'counts') #
raw_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
norm_counts <- LayerData(seu, assay = "RNA", layer = 'data') #
norm_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
```

Thereafter we use the HumanPrimaryCellAtlasData to find a reference dataset

```{r}

# 2. Get reference dataset
ref <- celldex::HumanPrimaryCellAtlasData()
unique(ref$label.main)
unique(ref$label.fine)

```

Since we have a PBMC dataset, there’s a lot of cell types in the reference dataset, like astrocytes, osteoblasts, hepatocytes.. that are not present in our dataset. This is why I’m going to only keep some blood cell types. However, you need to be careful when removing cell types from your reference, since they may be present in your dataset.

```{r}
# Subset to include only relevant cell types (CAREFUL!)
ref <- ref[,grepl('DC|B_cell|Neutrophils|T_cells|Monocyte|Erythroblast|
                 Macrophage|NK_cell|Platelets|Myelocyte', ref$label.main)]
unique(ref$label.main)
```

```{r}
# Call upon the reference database
# NOTE: reference database still undecided
ref <- as.SingleCellExperiment(obj, assay = "RNA")
```

```{r}
obj.counts <- GetAssayData(obj.cca, assay = "RNA", layer = 'counts')
```

```{r}
# Run SingleR
cell_type_pred <- SingleR(test = obj.counts,
                          ref = ref,
                          labels = ref$CellType,
                          de.method="wilcox")
```

```{r}
# Save the annotation results
obj.cca$cell_type_automatic <- cell_type_pred$pruned.labels[match(rownames(obj.cca@meta.data), rownames(cell_type_pred))]
```

```{r}
# Visualize
DimPlot(obj.cca, reduction = 'umap_cca', group.by = 'cell_type_automatic')
```
