---
title: "04_cell_type_annotation"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(multtest)
  library(metap)
  library(tibble)
  library(clustree)
  library(SingleR)
})

source("./paths.R")
#source("./functions.R")

EC <- readRDS(paste0(EC_seurats_path, "03_EC.rds"))
```

# Maunal Cell annotation

### Cell clustering

For cell type annotation, we will perform clustering first to gain insight into the local relationships of cell types.

```{r}
# Identify neighbors based on the CCA integration and perform clusering
EC <- FindNeighbors(EC,
                         dims = 1:15, 
                         reduction = "harmony_Integration", 
                         graph.name = "harmony_snn")

EC <- FindClusters(EC, 
                        resolution = c(0.4,0.6,0.8,1.0,1.2), 
                        algorithm = 4,
                        random.seed = 1,
                        graph.name = "harmony_snn")
```

### Clustering optimization

Clustering optimization is necessary to come to a clustering of cell types that we can confidently continue our analysis on. For optimization, we will test different resolutions (k) and determine the lowest resolution that provides a good clustering. To investigate this, we will create a clustree.

```{r}
clustree(EC, prefix = "harmony_snn_res.")
```

```{r}
# Set Idents to the resolution that was best
Idents(EC) <- "harmony_snn_res.1" 
```

#Tile plot

```{r}


df <- data.frame(
  cluster = EC@meta.data$harmony_snn_res.1,
  donor = EC@meta.data$donor
)

summary_df <- df %>%
  group_by(cluster, donor) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(fraction = count / sum(count))

```

```{r}
ggplot(summary_df, aes(x = donor, y = cluster, fill = fraction)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(fraction, 2)), color = "black", size = 3) + 
  scale_fill_gradient(low = "white", high = "red") +
  theme_minimal() +
  labs(x = "Donor", y = "Cluster", fill = "Fraction") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Cell type assignment

```{r}

# Definér hvilke donorer der er Disease
disease_donors <- c("B008", "B011")

# Opret ny metadata-kolonne
EC$Disease_status <- ifelse(
  EC$donor %in% disease_donors,
  "Disease",
  "Healthy"
)

# Tjek resultatet
table(EC$Disease_status, EC$donor)


```

### Feature plot

```{r}
# Example: CD14 is a gene typically expressed by monocytes.
FeaturePlot(EC, 
            reduction = "umap_harmony", 
            features = c("SMOC2"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE) 
```

**Manual marker cell type assignment**

It is also possible to utilize `FindConsveredMarkers()` and `FindMarkers()` for identifying genes that stand out in a cluster.

```{r}
# Compare one cluster to all the other clusters
# USE IF A CLUSTER IS STANDING OUT FROM OTHERS
clusterX_conserved_markers <- FindConservedMarkers(EC,
                                                   ident.1 = 7,
                                                   grouping.var = "donor",
                                                   only.pos = TRUE, min.pct = 0.25,
                                                   min.diff.pct = 0.05, 
                                                   logfc.threshold = 0.25)
```

```{r}
# Making feature grid plot for each of the features that are found

head(clusterX_conserved_markers)

head(rownames(clusterX_conserved_markers))

```

```{r}
# Find markers for all the clusters
for (cluster in levels(unique(Idents(EC) ))){
  p <- FindConservedMarkers(EC, ident.1 = cluster, grouping.var = "donor",
                              only.pos = TRUE, min.pct = 0.25,
                              min.diff.pct = 0.10, logfc.threshold = 0.25)
  print(cluster)
  print(head(p))
}
```

#Annotate the cell types from the found features

```{r}
# 1. Create a mapping dictionary named "cell_typ_dict" like below.
cell_typ_dict <- c( "1" = "CELL_TYPE_1", # replace these with real cell types.
                     "2" = "CELL_TYPE_2",
                     "3" = "CELL_TYPE_3",
                     "4" = "CELL_TYPE_4",
                     "5" = "Monocytes",
                     "6" = "M2 Macrophages",
                     "7" = "CELL_TYPE_7",
                     "8" = "CELL_TYPE_8",
                     "9" = "CELL_TYPE_9",
                     "10" = "CELL_TYPE_10",
                     "11" = "CELL_TYPE_11",
                     "12" = "Smooth Muscle Cells",
                     "13" = "Stem cells",
                     "14" = "CELL_TYPE_14",
                     "15" = "Goblet cells",
                     "16" = "CELL_TYPE_16",
                     "17" = "Epithelial Cells",
                     "18" = "Smooth Muscle Cells",
                     "19" = "Glandular Epithelial Cells",
                     "20" = "CELL_TYPE_20",
                     "21" = "Macrophages",
                     "22" = "Fibroblasts",
                     "23" = "Vascular Endothelial Cells",
                     "24" = "B cells",
                     "25" = "CELL_TYPE_25"
                    )
# 2. Map the clusters to the identified cell types and store them as a new column named "cell_type_manual".
EC$cell_type_manual <- unname(cell_typ_dict[as.character(Idents(EC))])

```

```{r}
clusters <- DimPlot(EC, reduction = 'umap_harmony', label = TRUE)
donor <- DimPlot(EC, reduction = 'umap_harmony', group.by = 'donor')
celltype <- DimPlot(EC, reduction = 'umap_harmony', group.by = 'cell_type_manual', label = TRUE) + NoLegend()
```

```{r}

saveRDS(EC, file = paste0(EC_seurats_path, "04_EC_manual.rds"))
```

# **Automatic cell type assignment**

following this tutorial: https://biostatsquid.com/singler-tutorial/

Initially, performing automatic cell type assignment due to not only the speed. but also now the quality of the reference database.

To run automatic cell type assignment the Seurat object has to be a SingleCellExperiment (SCE) object

First of all we need a cell annotation reference dataset

```{r}
#raw_counts <- LayerData(seu, assay = "RNA", layer = 'counts') #
#raw_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
#norm_counts <- LayerData(seu, assay = "RNA", layer = 'data') #
#norm_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
```

Thereafter we use the HumanPrimaryCellAtlasData to find a reference dataset

```{r}

# 2. Get reference dataset
#ref <- celldex::HumanPrimaryCellAtlasData()
#unique(ref$label.main)
#unique(ref$label.fine)

```

Since we have a PBMC dataset, there’s a lot of cell types in the reference dataset, like astrocytes, osteoblasts, hepatocytes.. that are not present in our dataset. This is why I’m going to only keep some blood cell types. However, you need to be careful when removing cell types from your reference, since they may be present in your dataset.

```{r}
# Subset to include only relevant cell types (CAREFUL!)
#ref <- ref[,grepl('DC|B_cell|Neutrophils|T_cells|Monocyte|Erythroblast|
#                 Macrophage|NK_cell|Platelets|Myelocyte', ref$label.main)]
#unique(ref$label.main)
```

#### Data import of reference material

```{r}
# # Load pathways
# rna_matrix   <- paste0(reference_data_path, "scipyref_matrix.mtx.gz")
# rna_features <- paste0(reference_data_path, "ref_features.tsv.gz")
# rna_barcodes <- paste0(reference_data_path, "ref_barcodes.tsv.gz")
# 
# rna_features_data <- read.csv(paste0(reference_data_path, "ref_features2.tsv.gz"), sep="\t")
# rna_features_data <- rna_features_data[, c("ENSG00000243485", "MIR1302.2HG", "Gene.Expression")]
# 
# write.table(rna_features_data, 
#             rna_features,
#             row.names=FALSE,
#             sep ="\t") # Remove index
```

```{r}
# Create Seurat object
# counts <- ReadMtx(
#   mtx = rna_matrix,
#   features = rna_features,
#   cells = rna_barcodes
# )
# 
# REF <- CreateSeuratObject(
#   counts = counts,
#   project = "reference"
# )
# 
# saveRDS(REF, file = paste0(reference_data_path, "ref.rds"))
```

```{r}
#REF <- readRDS(paste0(reference_data_path, "ref.rds"))
```

```{r}
# # Reference meta data
# metadata <- read.csv(paste0(reference_data_path, "all_ref_data.csv"), sep=";")
# 
# # Check the first few rows to ensure 'cell_type' is present
# head(metadata[, "cell_type", drop = FALSE])
```

```{r}
# # Add cell type
# REF <- AddMetaData(
#   object = REF,
#   metadata = metadata[, "cell_type"],
#   col.name = "cell_type"
# )
# 
# # Add tissue
# REF <- AddMetaData(
#   object = REF,
#   metadata = metadata[, "tissue"],
#   col.name = "tissue"
# )
# 
# # Add highlevel cell type category
# REF <- AddMetaData(
#   object = REF,
#   metadata = metadata[, "category"],
#   col.name = "cell_category"
# )
# 
# # Add disease
# REF <- AddMetaData(
#   object = REF,
#   metadata = metadata[, "disease"],
#   col.name = "disease_state"
# )
# 
# # Add diagnosis. Options = (fetal, healthy adult, pediatric crohn's disease, pediatric healthy)
# REF <- AddMetaData(
#   object = REF,
#   metadata = metadata[, "Diagnosis"],
#   col.name = "ref_diagnosis"
# )
```

#### Subset reference data

We want to first subset based on 'tissue'. Here, we would like to only work with values 'ascending colon' and 'transverse colon'.

```{r}
# # Subset based on tissue
# REF.filtered <- subset(REF,
#                         tissue %in% c("ascending colon", "transverse colon"))
# 
# # Subset based on disease_state -> we would only like to keep 'normal'
# REF.filtered <- subset(REF.filtered,
#                        disease_state == "normal")
# 
# # Subset based on ref_diagnosis -> we would like to discard "Pediatric Crohn Disease"
# REF.filtered <- subset(REF.filtered,
#                        ref_diagnosis != "Pediatric Crohn Disease")
```

Normalize the data (get logcounts)

```{r}
# #| warning: false
# #| message: false
# 
# REF.filtered <- NormalizeData(REF.filtered)
# REF.filtered <- FindVariableFeatures(REF.filtered)
# REF.filtered <- ScaleData(REF.filtered)
```

Save subsetted reference data

```{r}
# saveRDS(REF.filtered, file = paste0(reference_data_path, "ref_subset2.rds"))
```

#### Run SingleR

```{r}
# Load reference data
ref <- readRDS(paste0(reference_data_path, "ref_subset2.rds"))
```

```{r}
# Work with the reference database as a SCE object
ref <- as.SingleCellExperiment(ref, assay = "RNA")
```

```{r}
EC_counts <- GetAssayData(EC, assay = "RNA", layer = 'counts')
```

```{r}
# Run SingleR
cell_category_pred <- SingleR(test = EC_counts,
                          ref = ref,
                          labels = ref$cell_category, #cell_type for more specific view
                          de.method="wilcox")
```

```{r}
saveRDS(cell_category_pred, paste0(reference_data_path, "cell_category_pred.singler.rds"))
```

```{r}
# Save the annotation results
EC$cell_type_automatic <- cell_category_pred$pruned.labels[match(rownames(EC@meta.data), 
                                                             rownames(cell_category_pred))]
```

```{r}
# Visualize
DimPlot(EC, reduction = 'umap_harmony', group.by = 'cell_type_automatic', pt.size = 0.1) +
  ggtitle("Cell category annotation on UMAP harmony") +
  theme(legend.text = element_text(size = 8), 
        plot.title = element_text(hjust = 0.5, face = "bold")) + 
  guides(fill = guide_legend(ncol = 3))         # Force the legend into multiple columns
```

The more detailed cell type was previously run and stored as:

```{r}
cell_type_pred <- readRDS(paste0(reference_data_path, "cell_type_pred.singler.rds"))
```

Load and redraw the umap embedding space, but with filtered annotations. All of them will crowd the plotting space (we're anyways only interested in intestine related cells).

```{r}
head(cell_type_pred)
```

metadata can be used to lookup for cell_category == epithelial cells; corresponding cell_type labels

```{r}
metadata <- read.csv(paste0(reference_data_path, "all_ref_data.csv"), sep=";")
head(metadata)
```

```{r}
metadata_intestine <- metadata %>% 
  mutate(intestine = case_when(
    category == "Epithelial" ~ TRUE,
    .default = FALSE)) %>% 
  select(cell_type, intestine) 

metadata_intestine
```

```{r}
library(dplyr)

# 1. Get a list of the cell types that are actually "intestine"
true_intestine_types <- metadata_intestine %>%
  filter(intestine == TRUE) %>%
  pull(cell_type)

# 2. Filter the prediction dataframe
filtered_preds <- cell_type_pred %>%
  as.data.frame() %>% # Ensure it's a dataframe if it's a Seurat/SingleR object
  filter(labels %in% true_intestine_types)

head(filtered_preds)
```

```{r}
# Save the annotation results
EC$cell_type_automatic <- filtered_preds$pruned.labels[match(rownames(EC@meta.data), 
                                                             rownames(filtered_preds))]
```

```{r}
# Visualize
DimPlot(EC, reduction = 'umap_harmony', group.by = 'cell_type_automatic', pt.size = 0.05, alpha=0.4) +
  ggtitle("Cell type annotation on UMAP harmony") +
  theme(legend.text = element_text(size = 8), 
        plot.title = element_text(hjust = 0.5, face = "bold")) + 
  guides(fill = guide_legend(ncol = 3))         # Force the legend into multiple columns
```
