---
title: "04_cell_type_annotation"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(multtest)
library(metap)
library(tibble)
library(clustree)
library(SingleR)
})
```

### Cell clustering

For cell type annotation, we will perform clustering first to gain insight into the local relationships of cell types.

```{r}
# Identify neighbors based on the CCA integration and perform clusering
# REMEMBER TO UPDATE PCS
obj.cca <- FindNeighbors(obj,
                         dims = 1:14, 
                         reduction = "CCA_Integration", 
                         graph.name = "CCA_snn")

obj.cca <- FindClusters(obj.cca, 
                        resolution = c(0.4, 0.6, 0.8, 1.0, 1.2), 
                        algorithm = 4, 
                        graph.name = "CCA_snn")
```

### Clustering optimization

Clustering optimization is necessary to come to a clustering of cell types that we can confidently continue our analysis on. For optimization, we will test different resolutions (k) and determine the lowest resolution that provides a good clustering. To investigate this, we will create a clustree.

```{r}
clustree(obj.cca, prefix = "CCA_snn_res.")
```

```{r}
# Set Idents to the resolution that was best
Idents(obj.cca) <- "CCA_snn_res.X.X" 
```

### Cell type assignment

Initially, performing automatic cell type assignment due to not only the speed. but also now the quality of the reference database.

To run automatic cell type assignment the Seurat object has to be a SingleCellExperiment (SCE) object

```{r}
# Call upon the reference database
# NOTE: reference database still undecided
ref <- as.SingleCellExperiment(obj, assay = "RNA")
```

```{r}
obj.counts <- GetAssayData(obj.cca, assay = "RNA", layer = 'counts')
```

```{r}
# Run SingleR
cell_type_pred <- SingleR(test = obj.counts,
                          ref = ref,
                          labels = ref$CellType,
                          de.method="wilcox")
```

```{r}
# Save the annotation results
obj.cca$cell_type_automatic <- cell_type_pred$pruned.labels[match(rownames(obj.cca@meta.data), rownames(cell_type_pred))]
```

```{r}
# Visualize
DimPlot(obj.cca, reduction = 'umap_cca', group.by = 'cell_type_automatic')
```
