---
title: "04_cell_type_annotation"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(multtest)
library(metap)
library(tibble)
library(clustree)
library(SingleR)
})

source("./paths.R")
#source("./functions.R")
```

### Cell clustering

For cell type annotation, we will perform clustering first to gain insight into the local relationships of cell types.

```{r}
# Identify neighbors based on the CCA integration and perform clusering
# REMEMBER TO UPDATE PCS
obj.cca <- FindNeighbors(obj,
                         dims = 1:14, 
                         reduction = "CCA_Integration", 
                         graph.name = "CCA_snn")

obj.cca <- FindClusters(obj.cca, 
                        resolution = c(0.4, 0.6, 0.8, 1.0, 1.2), 
                        algorithm = 4, 
                        graph.name = "CCA_snn")
```

### Clustering optimization

Clustering optimization is necessary to come to a clustering of cell types that we can confidently continue our analysis on. For optimization, we will test different resolutions (k) and determine the lowest resolution that provides a good clustering. To investigate this, we will create a clustree.

```{r}
clustree(obj.cca, prefix = "CCA_snn_res.")
```

```{r}
# Set Idents to the resolution that was best
Idents(obj.cca) <- "CCA_snn_res.X.X" 
```

### Cell type assignment

**Manual marker cell type assignment**

Cell type assignment by manual markers requires prior knowledge about characteristic markers for the cells that we are expecting in the data. The expression of the marker can be visualized with the `FeaturePlot()`

```{r}
# Visualize gene expression
FeaturePlot(obj.cca, 
            reduction = "umap_cca", 
            features = c("XXX"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

It is also possible to utilize `FindConsveredMarkers()` and `FindMarkers()` for identifying genes that stand out in a cluster.

```{r}
# Compare one cluster to all the other clusters
# USE IF A CLUSTER IS STANDING OUT FROM OTHERS
clusterX_conserved_markers <- FindConservedMarkers(obj.cca,
                                                   ident.1 = 10,
                                                   grouping.var = "condition",
                                                   only.pos = TRUE, min.pct = 0.25,
                                                   min.diff.pct = 0.25, 
                                                   logfc.threshold = 0.25)
```

```{r}
# Compare one cluster to another
clusterX_clusterY_markers <- FindMarkers(obj.cca,
                                         ident.1 = X,
                                         ident.2 = Y,
                                         grouping.var = "condition",
                                         only.pos = TRUE, 
                                         min.pct = 0.25,
                                         min.diff.pct = 0.25, 
                                         logfc.threshold = 0.25)
```

**Automatic cell type assignment**

following this tutorial: https://biostatsquid.com/singler-tutorial/

Initially, performing automatic cell type assignment due to not only the speed. but also now the quality of the reference database.

To run automatic cell type assignment the Seurat object has to be a SingleCellExperiment (SCE) object

First of all we need a cell annotation reference dataset

```{r}
raw_counts <- LayerData(seu, assay = "RNA", layer = 'counts') #
raw_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
norm_counts <- LayerData(seu, assay = "RNA", layer = 'data') #
norm_counts[c('VIM', 'BCL2', 'TP53', 'CD4'),1:5]
```

Thereafter we use the HumanPrimaryCellAtlasData to find a reference dataset

```{r}

# 2. Get reference dataset
ref <- celldex::HumanPrimaryCellAtlasData()
unique(ref$label.main)
unique(ref$label.fine)

```

Since we have a PBMC dataset, there’s a lot of cell types in the reference dataset, like astrocytes, osteoblasts, hepatocytes.. that are not present in our dataset. This is why I’m going to only keep some blood cell types. However, you need to be careful when removing cell types from your reference, since they may be present in your dataset.

```{r}
# Subset to include only relevant cell types (CAREFUL!)
ref <- ref[,grepl('DC|B_cell|Neutrophils|T_cells|Monocyte|Erythroblast|
                 Macrophage|NK_cell|Platelets|Myelocyte', ref$label.main)]
unique(ref$label.main)
```

```{r}
# Call upon the reference database
# NOTE: reference database still undecided
ref <- as.SingleCellExperiment(obj, assay = "RNA")
```

```{r}
obj.counts <- GetAssayData(obj.cca, assay = "RNA", layer = 'counts')
```

```{r}
# Run SingleR
cell_type_pred <- SingleR(test = obj.counts,
                          ref = ref,
                          labels = ref$CellType,
                          de.method="wilcox")
```

```{r}
# Save the annotation results
obj.cca$cell_type_automatic <- cell_type_pred$pruned.labels[match(rownames(obj.cca@meta.data), rownames(cell_type_pred))]
```

```{r}
# Visualize
DimPlot(obj.cca, reduction = 'umap_cca', group.by = 'cell_type_automatic')
```
