---
title: "07_differential_gene_expression"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(multtest)
  library(metap)
  library(tibble)
  library(DESeq2)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(jsonlite)
})

source("./paths.R")
```



Load in data

```{r}
EC <- readRDS(paste0(EC_seurats_path, "04_EC.rds"))
```


Creating disase_sample

```{r}
EC@meta.data$sample_disease <- paste(
  EC@meta.data$donor,
  EC@meta.data$Disease_status,
  sep = "_"
)
```


### DESeq2


```{r}

# Pseudobulk data
# Aggregate counts to donor level
# REMEMBER to doublecheck if we are grouping by cell type and donor
counts <- AggregateExpression(EC, 
                              group.by = c("cell_type_automatic", "sample_disease"),
                              assays =  "RNA",
                              return.seurat = FALSE)

counts <- counts$RNA

# transpose
counts.t <- t(counts)

# convert to data.frame
counts.t <- as.data.frame(counts.t)

# get values where to split
splitRows <- gsub('_.*', '', rownames(counts.t))

# split data.frame
cts.split <- split.data.frame(counts.t,
                              f = factor(splitRows))
# fix colnames and transpose
cts.split.modified <- lapply(cts.split, function(x){
    rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x)) #whatever comes after the underscore in the row names will be retained in the final result.
    t(x)
})
```

```{r}
# REMEMBER to replace 'cell_type' with e.g. 'CD14+ Monocyte'

# EXAMPLE:
# counts_monocytes <- cts.split.modified$`CD14+ Monocyte`

# SUGGESTION: update the name 'counts_cell_type'
counts_epithelial <- cts.split.modified$`colon epithelial cell`
```

```{r}
# Define the colnames to be the colnames of the cell_type count matrix
# REMEMEBER to update 'counts_cell_type' these are all from the same study and therefore only have to be done once.
colData <- data.frame(samples = colnames(counts_epithelial))

colData <- colData %>%
  dplyr::mutate(condition = ifelse(grepl("Disease", samples), 'Disease', 'Healthy')) %>%
  column_to_rownames(var = 'samples')

```

Create a DESeq2 object to use for the function `DESeqDataSetFromMatrix()`

```{r}
# Create DESeq2 object   
# REMEMEBER to update 'counts_cell_type'
dds_epithelial <- DESeqDataSetFromMatrix(countData = counts_epithelial,
                              colData = colData,
                              design = ~ condition)


```

```{r}
# Remove rows (genes) that have fewer than 10 reads.
keep <- rowSums(counts(dds_epithelial)) >=10
dds_epithelial <- dds_epithelial[keep,]
```

```{r}
# Run DESeq2
dds_epithelial <- DESeq(dds_epithelial)

# Save the resuslts
res_epithelial <- results(dds_epithelial, name = "condition_Healthy_vs_Disease")
summary(res_epithelial)
```

We can now identify the top 20 differentially expressed genes in cell type X in sick donors compared to control donors. You can either sort by p-adjusted or log2FoldChange. padj reflects the significance of results while log2FoldChange reflects the "power" of the difference between the conditions (NOTE: account for negative and positive value by looking at the absolute values).

```{r}
# Set thresholds
padj_cutoff <- 0.05

# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl_epithelial <- res_epithelial %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

# Subset the significant results
sig_res_epithelial <- dplyr::filter(res_tbl_epithelial, padj < padj_cutoff)

```

```{r}
# Order results by padj values.
top20_sig_genes_epithelial <- sig_res_epithelial %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_epithelial)
```

```{r}
# Order results by log2FoldChange values.
top20_sig_genes_epithelial <- sig_res_epithelial %>%
  dplyr::arrange(log2FoldChange) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_epithelial)
```


```{r}
# REMEMBER to replace 'cell_type' with e.g. 'CD14+ Monocyte'

# EXAMPLE:
# counts_monocytes <- cts.split.modified$`CD14+ Monocyte`

# SUGGESTION: update the name 'counts_cell_type'
counts_glial_cells <- cts.split.modified$`glial cell`
```

```{r}
# # Define the colnames to be the colnames of the cell_type count matrix
# # REMEMEBER to update 'counts_cell_type'
# colData <- data.frame(samples = colnames(counts_glial_cells))
# 
# colData <- colData %>%
#   dplyr::mutate(condition = ifelse(grepl("Disease", samples), 'Disease', 'Healthy')) %>%
#   column_to_rownames(var = 'samples')

```

Create a DESeq2 object to use for the function `DESeqDataSetFromMatrix()`

```{r}
# Create DESeq2 object   
# REMEMEBER to update 'counts_cell_type'
dds_glial <- DESeqDataSetFromMatrix(countData = counts_glial_cells,
                              colData = colData,
                              design = ~ condition)


```

```{r}
# Remove rows (genes) that have fewer than 10 reads.
keep <- rowSums(counts(dds_glial)) >=10
dds_glial <- dds_glial[keep,]
```

```{r}
# Run DESeq2
dds_glial <- DESeq(dds_glial)

# Save the resuslts
res_glial <- results(dds_glial, name = "condition_Healthy_vs_Disease")
summary(res_glial)
```
There are no significantly differentially expressed glial cell genes


```{r}
# REMEMBER to replace 'cell_type' with e.g. 'CD14+ Monocyte'

# EXAMPLE:
# counts_monocytes <- cts.split.modified$`CD14+ Monocyte`

# SUGGESTION: update the name 'counts_cell_type'
counts_goblet_cell <- cts.split.modified$`intestine goblet cell`
```

```{r}
# # Define the colnames to be the colnames of the cell_type count matrix
# # REMEMEBER to update 'counts_cell_type'
# colData <- data.frame(samples = colnames(counts_goblet_cell))
# 
# colData <- colData %>%
#   dplyr::mutate(condition = ifelse(grepl("Disease", samples), 'Disease', 'Healthy')) %>%
#   column_to_rownames(var = 'samples')

```

Create a DESeq2 object to use for the function `DESeqDataSetFromMatrix()`

```{r}
# Create DESeq2 object   
# REMEMEBER to update 'counts_cell_type'
dds_goblet_cell <- DESeqDataSetFromMatrix(countData = counts_goblet_cell,
                              colData = colData,
                              design = ~ condition)


```

```{r}
# Remove rows (genes) that have fewer than 10 reads.
keep <- rowSums(counts(dds_goblet_cell)) >=10
dds_goblet_cell <- dds_goblet_cell[keep,]
```

```{r}
# Run DESeq2
dds_goblet_cell <- DESeq(dds_goblet_cell)

# Save the resuslts
res_goblet_cell <- results(dds_goblet_cell, name = "condition_Healthy_vs_Disease")
summary(res_goblet_cell)
```

We can now identify the top 20 differentially expressed genes in cell type X in sick donors compared to control donors. You can either sort by p-adjusted or log2FoldChange. padj reflects the significance of results while log2FoldChange reflects the "power" of the difference between the conditions (NOTE: account for negative and positive value by looking at the absolute values).

```{r}
# Set thresholds
padj_cutoff <- 0.05

# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl_goblet_cell <- res_goblet_cell %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

# Subset the significant results
sig_res_goblet_cell <- dplyr::filter(res_tbl_goblet_cell, padj < padj_cutoff)
```

```{r}
# Order results by padj values.
top20_sig_genes_goblet_cell <- sig_res_goblet_cell %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_goblet_cell)
```

```{r}
# Order results by log2FoldChange values.
top20_sig_genes_goblet_cell <- sig_res_goblet_cell %>%
  dplyr::arrange(log2FoldChange) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_goblet_cell)
```



```{r}
# REMEMBER to replace 'cell_type' with e.g. 'CD14+ Monocyte'

# EXAMPLE:
# counts_monocytes <- cts.split.modified$`CD14+ Monocyte`

# SUGGESTION: update the name 'counts_cell_type'
counts_macrophage <- cts.split.modified$`macrophage`
```

```{r}
# # Define the colnames to be the colnames of the cell_type count matrix
# # REMEMEBER to update 'counts_cell_type'
# colData <- data.frame(samples = colnames(counts_macrophage))
# 
# colData <- colData %>%
#   dplyr::mutate(condition = ifelse(grepl("Disease", samples), 'Disease', 'Healthy')) %>%
#   column_to_rownames(var = 'samples')

```

Create a DESeq2 object to use for the function `DESeqDataSetFromMatrix()`

```{r}
# Create DESeq2 object   
# REMEMEBER to update 'counts_cell_type'
dds_macrophage <- DESeqDataSetFromMatrix(countData = counts_macrophage,
                              colData = colData,
                              design = ~ condition)


```

```{r}
# Remove rows (genes) that have fewer than 10 reads.
keep <- rowSums(counts(dds_macrophage)) >=10
dds_macrophage <- dds_macrophage[keep,]
```

```{r}
# Run DESeq2
dds_macrophage <- DESeq(dds_macrophage)

# Save the resuslts
res_macrophage <- results(dds_macrophage, name = "condition_Healthy_vs_Disease")
summary(res_macrophage)
```

We can now identify the top 20 differentially expressed genes in cell type X in sick donors compared to control donors. You can either sort by p-adjusted or log2FoldChange. padj reflects the significance of results while log2FoldChange reflects the "power" of the difference between the conditions (NOTE: account for negative and positive value by looking at the absolute values).

```{r}
# Set thresholds
padj_cutoff <- 0.05

# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl_macrophage <- res_macrophage %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

# Subset the significant results
sig_res_macrophage <- dplyr::filter(res_tbl_macrophage, padj < padj_cutoff)
```

```{r}
# Order results by padj values.
top20_sig_genes_macrophage <- sig_res_macrophage %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_macrophage)
```

```{r}
# Order results by log2FoldChange values.
top20_sig_genes_macrophage <- sig_res_macrophage %>%
  dplyr::arrange(log2FoldChange) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes_macrophage)
```



### fgsea

Depending on the library different gene-sets can be obtained. For this project, we will utilize the gene-sets from the exercises


loading in the gene set data

```{r}

#For Geneset Ontology
geneset_ontology <- fromJSON("/home/projects/exam_2026_22102/group2/gsea_dataset/c5.all.v2025.1.Hs.txt")

#For Diverse Pathways
gsList <- readRDS("/home/projects/22102_single_cell/day5/gene_sets.Rds")


#For cell types
gene_sets <- fromJSON("/home/projects/exam_2026_22102/group2/gsea_dataset/c8.all.v2025.1.Hs.txt")



```


here we create the gene set list that we want to use!!

```{r}
# Extract only the genes into a named list
gene_set_list <- lapply(gene_sets, function(x) x$geneSymbols)
names(gene_set_list) <- names(gene_sets)


```



```{r}
# Extract only the genes into a named list
geneset_ontology <- lapply(geneset_ontology, function(x) x$geneSymbols)
names(geneset_ontology) <- names(geneset_ontology)


```



#####-----------------Epithelial Cells--------------------------------------###



###-------------------------------For KEGG----------------------------------###


```{r}

res_epithelial$genes <- rownames(res_epithelial)
#Add EntrezID for the genes
library(org.Hs.eg.db)
Entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                    key=res_epithelial$genes, 
                                    columns="ENTREZID",
                                    keytype="SYMBOL")

Entrez <- as_tibble(Entrez)
res_epithelial <- as.data.frame(res_epithelial)
res_epithelial <- inner_join(res_epithelial, Entrez, by=c("genes"="SYMBOL"))
#Remove those genes that do not have an entrezID
res_epithelial <- filter(res_epithelial, !is.na(ENTREZID))


```


Ensure no p avlues are na or 0( Check this!!)

```{r}
res_epithelial <- res_epithelial %>%
  filter(!is.na(log2FoldChange), !is.na(pvalue), pvalue > 0)

```

There are two duplikate genes: "TEC"  "MMD2". I chose only so retain the one with.

```{r}
res_epithelial <- res_epithelial %>%
  filter(!genes %in% c("TEC", "MMD2"))
```



```{r}


# Create a ranked gene list. 
# In the exercises they use:    -log10(pvalue) * sign(log2FoldChange)

# Create new metric 'stat_sig'
res_epithelial$stat_sig <- -log10(res_epithelial$pvalue) * sign(res_epithelial$log2FoldChange)
rankData_epithelial <- res_epithelial$stat_sig
names(rankData_epithelial) <- res_epithelial$genes

```



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gsList, # gene-set reference list
                  stats=rankData_epithelial)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```





We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

g1 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark pathways for epithelial cells from KEGG GSEA",
    width = 70
  )) + 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(g1)
```


###-----------------------------For cell types------------------------------###



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gene_set_list, # gene-set reference list
                  stats=rankData_epithelial)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```


We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

g2 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark pathways for epithelial cells from cell-type signature geneset GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(g2)
```


#-------------------------Gene Ontology-------------------------------------##



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=geneset_ontology, # gene-set reference list
                  stats=rankData_epithelial)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```



We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

g3 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark pathways for epithelial cells from Geneset ontology GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(g3)
```




#####-----------------Goblet Cells--------------------------------------###



###-------------------------------For KEGG----------------------------------###


```{r}

res_goblet_cell$genes <- rownames(res_goblet_cell)
#Add EntrezID for the genes
library(org.Hs.eg.db)
Entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                    key=res_goblet_cell$genes, 
                                    columns="ENTREZID",
                                    keytype="SYMBOL")

Entrez <- as_tibble(Entrez)
res_goblet_cell <- as.data.frame(res_goblet_cell)
res_goblet_cell <- inner_join(res_goblet_cell, Entrez, by=c("genes"="SYMBOL"))
#Remove those genes that do not have an entrezID
res_goblet_cell <- filter(res_goblet_cell, !is.na(ENTREZID))


```


Ensure no p avlues are na or 0( two p-values are NA - they are removed)

```{r}
res_goblet_cell <- res_goblet_cell %>%
  filter(!is.na(log2FoldChange), !is.na(pvalue), pvalue > 0)

```

There are two duplikate genes: "TEC"  "MMD2". I chose only so retain the one with.

```{r}
res_goblet_cell <- res_goblet_cell %>%
  filter(!genes %in% c("TEC", "MMD2"))
```


```{r}
any(duplicated(rownames(res_goblet_cell)))
```



```{r}


# Create a ranked gene list. 
# In the exercises they use:    -log10(pvalue) * sign(log2FoldChange)

# Create new metric 'stat_sig'
res_goblet_cell$stat_sig <- -log10(res_goblet_cell$pvalue) * sign(res_goblet_cell$log2FoldChange)
rankData_goblet_cell <- res_goblet_cell$stat_sig
names(rankData_goblet_cell) <- res_goblet_cell$genes

```



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gsList, # gene-set reference list
                  stats=rankData_goblet_cell)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```





We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

h1 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark Pathways for Goblet Cells from KEGG GSEA",
    width = 70
  )) + 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(h1)
```


###-----------------------------For cell types------------------------------###



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gene_set_list, # gene-set reference list
                  stats=rankData_goblet_cell)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```


We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

h2 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark Pathways for Goblet Cells from Cell-type Signature Geneset GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(h2)
```


#-------------------------Gene Ontology-------------------------------------##



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=geneset_ontology, # gene-set reference list
                  stats=rankData_goblet_cell)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```



We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

h3 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark pathways for Goblet Cells from Geneset ontology GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(h3)
```










#####--------------------Macrophage cells--------------------------------------###



###-------------------------------For KEGG----------------------------------###


```{r}

res_macrophage$genes <- rownames(res_macrophage)
#Add EntrezID for the genes
library(org.Hs.eg.db)
Entrez <- AnnotationDbi::select(org.Hs.eg.db,
                                    key=res_macrophage$genes, 
                                    columns="ENTREZID",
                                    keytype="SYMBOL")

Entrez <- as_tibble(Entrez)
res_macrophage <- as.data.frame(res_macrophage)
res_macrophage <- inner_join(res_macrophage, Entrez, by=c("genes"="SYMBOL"))
#Remove those genes that do not have an entrezID
res_macrophage <- filter(res_macrophage, !is.na(ENTREZID))


```


Ensure no p avlues are na or 0( two p-values are NA - they are removed)

```{r}
res_macrophage <- res_macrophage %>%
  filter(!is.na(log2FoldChange), !is.na(pvalue), pvalue > 0)

```

There are two duplikate genes: "TEC"  "MMD2". I chose only so retain the one with.

```{r}
res_macrophage <- res_macrophage %>%
  filter(!genes %in% c("TEC", "MMD2"))
```


```{r}
any(duplicated(rownames(res_macrophage)))
```



```{r}


# Create a ranked gene list. 
# In the exercises they use:    -log10(pvalue) * sign(log2FoldChange)

# Create new metric 'stat_sig'
res_macrophage$stat_sig <- -log10(res_macrophage$pvalue) * sign(res_macrophage$log2FoldChange)
rankData_macrophage <- res_macrophage$stat_sig
names(rankData_macrophage) <- res_macrophage$genes

```



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gsList, # gene-set reference list
                  stats=rankData_macrophage)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```





We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

j1 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark Pathways for Macrophage Cells from KEGG GSEA",
    width = 70
  )) + 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(j1)
```


###-----------------------------For cell types------------------------------###



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=gene_set_list, # gene-set reference list
                  stats=rankData_macrophage)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```


We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

j2 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark Pathways for Macrophage Cells from Cell-type Signature Geneset GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(j2)
```


#-------------------------Gene Ontology-------------------------------------##



```{r}

# Run fgsea algorithm
fgseaRes <- fgsea(pathways=geneset_ontology, # gene-set reference list
                  stats=rankData_macrophage)  # ranked gene list

# Tidy the results
fgseaResTidy<- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

head(fgseaResTidy)

```



We can then plot the top 10 up-/down-regulated pathways enriched in cell type X for the sick donors

```{r}
# Sort the data frame based on NES in descending order to get top 10 pathways
top_10 <- fgseaResTidy %>%
  arrange(desc(NES)) %>%
  head(10)

# Sort the data frame based on NES in ascending order to get bottom 10 pathways
bottom_10 <- fgseaResTidy %>%
  arrange(NES) %>%
  head(10)

# Combine top 10 and bottom 10 pathways
top_bottom_10 <- rbind(top_10, bottom_10)

top_bottom_10 <- top_bottom_10 %>%
  mutate(signif = factor(padj < 0.05, levels = c(TRUE, FALSE)))

j3 <- ggplot(top_bottom_10, 
            aes(reorder(pathway, NES), NES)) +
  
  geom_col(aes(fill = signif)) +
  coord_flip() +
  labs(x="Pathway", 
       y="Normalized Enrichment Score",
       title = stringr::str_wrap(
    "Top 10 and Bottom 10 Hallmark Pathways for Macrophage Cells from Geneset Ontology GSEA",
    width = 70
  ))+ 
  scale_fill_manual(values = c("TRUE" = "#00BFC4", "FALSE" = "#F8766D"),
                  name = "padj < 0.05")+
  theme_minimal()

print(j3)
```




#done





