---
title: "07_differential_gene_expression"
format: html
---

```{r}
# Load libraries
suppressPackageStartupMessages({
  library(Seurat)
  library(ggplot2)
  library(patchwork)
  library(dplyr)
  library(tidyr)
  library(multtest)
  library(metap)
  library(tibble)
  library(DESeq2)
})

source("./paths.R")
```

### DESeq2

```{r}
# Pseudobulk data
# Aggregate counts to donor level
# REMEMBER to doublecheck if we are grouping by cell type and donor
counts <- AggregateExpression(obj, 
                              group.by = c("cell_type_manual", "donor"),
                              assays =  "RNA",
                              return.seurat = FALSE)

# Transpose
counts.t <- t(counts)

# Convert to data.frame
counts.t <- as.data.frame(counts.t)

# Get values where to split
# EXAMPLE: from 'Bcell_sample1' to 'Bcell'
splitRows <- gsub('_.*', '', rownames(counts.t))
```

```{r}
# Split data.frame
cts.split <- split.data.frame(counts.t,
                              f = factor(splitRows))
# Fix colnames and transpose
cts.split.modified <- lapply(cts.split, function(x){
    rownames(x) <- gsub('.*_(.*)', '\\1', rownames(x)) 
    # EXAMPLE: from 'Bcell_sample1' to 'sample1'
    t(x)
})
```

We then "subset" the data looking only at a target cell type and its count matrix.

```{r}
# REMEMBER to replace 'cell_type' with e.g. 'CD14+ Monocyte'

# EXAMPLE:
# counts_monocytes <- cts.split.modified$`CD14+ Monocyte`

# SUGGESTION: update the name 'counts_cell_type'
counts_cell_type <- cts.split.modified$`cell_type`
```

```{r}
# Define the colnames to be the colnames of the cell_type count matrix
# REMEMEBER to update 'counts_cell_type'
colData <- data.frame(samples = colnames(counts_cell_type))

# Grep only hypertension donors and assign 'Sick'-label
colData <- colData %>%
  # Replace 'hypertension' with the string that defines the donor as sick
  # in the 'samples'-column. 
  dplyr::mutate(condition = ifelse(grepl('hypertension', samples), 
                                   'Sick', 
                                   'Normal')) %>%
  
  column_to_rownames(var = 'samples')
```

Create a DESeq2 object to use for the function `DESeqDataSetFromMatrix()`

```{r}
# Create DESeq2 object   
# REMEMEBER to update 'counts_cell_type'
dds <- DESeqDataSetFromMatrix(countData = counts_cell_typee,
                              colData = colData,
                              design = ~ condition)
```

```{r}
# Remove rows (genes) that have fewer than 10 reads.
keep <- rowSums(counts(dds)) >=10
dds <- dds[keep,]
```

```{r}
# Run DESeq2
dds <- DESeq(dds)

# Save the resuslts
res <- result(dds, name = "condition_Sick_vs_Control")
summary(res)
```

We can now identify the top 20 differentially expressed genes in cell type X in sick donors compared to control donors. You can either sort by p-adjusted or log2FoldChange. padj reflects the significance of results while log2FoldChange reflects the "power" of the difference between the conditions (NOTE: account for negative and positive value by looking at the absolute values).

```{r}
# Set thresholds
padj_cutoff <- 0.05

# Turn the DESeq2 results object into a tibble for use with tidyverse functions
res_tbl <- res %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

# Subset the significant results
sig_res <- dplyr::filter(res_tbl, padj < padj_cutoff)
```

```{r}
# Order results by padj values.
top20_sig_genes <- sig_res %>%
  dplyr::arrange(padj) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes)
```

```{r}
# Order results by log2FoldChange values.
top20_sig_genes <- sig_res %>%
  dplyr::arrange(log2FoldChange) %>%
  dplyr::pull(gene) %>%
  head(n=20)

print(top20_sig_genes)
```

### fgsea
