---
title: "07_gene_regulation_interation"
format: html
---

## Loading data and libraries

```{r}
suppressPackageStartupMessages({
library(Signac)
library(Seurat)
library(ggplot2)
library(dplyr)
})

source("./paths.R")
#source("./functions.R")
```

```{r}
EC <- readRDS(paste0(EC_seurats_path, "04_EC.rds"))
print(EC)
```

## QC of ATAC

```{r}
DefaultAssay(EC) <- "ATAC" 

EC$orig_cluster_ident <- Idents(EC)  # backup current cluster identities
EC$all <- "all cells"
Idents(EC) <- "all"

EC@meta.data$subset_nFeature_ATAC <- colSums(GetAssayData(EC, assay = "ATAC", layer = "counts") > 0)


VlnPlot(EC, features = c("nCount_ATAC"), ncol = 2,
  log = TRUE, pt.size = 0) + NoLegend()
```

```{r}

atac_cells <- WhichCells(EC, expression = nCount_ATAC > 500 & nCount_ATAC < 90000)

# intersecting 
EC <- subset(EC, cells= atac_cells)

```

## Data processing

```{r}
DefaultAssay(EC) <- "ATAC"

EC <- RunTFIDF(EC)
EC <- FindTopFeatures(EC, min.cutoff="q0")
EC <- RunSVD(EC)
```

Checking dimensions - we want to remove first one as it often explains technical variance

```{r}
DepthCor(EC)
```

```{r}
EC <- RunUMAP(EC, reduction = "lsi", dims = 2:30, reduction.name = "umap.atac", reduction.key = "atacUMAP_")
```

```{r}
Idents(EC) <- 'donor'

DimPlot(EC, reduction='umap.atac') + ggtitle("ATAC non-integrated")
```

## Data integration - vertical

```{r}

EC <- FindMultiModalNeighbors(EC, reduction.list = list("harmony_Integration", "lsi"), dims.list = list(1:15, 2:30))

```

```{r}
EC <- RunUMAP(EC, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
```

```{r}
EC <- FindClusters(EC, graph.name = "wsnn", algorithm = 4, verbose = FALSE)
```

```{r}

p1 <- DimPlot(EC, reduction='umap_harmony') + ggtitle("RNA")
p2 <- DimPlot(EC, reduction='umap.atac') + ggtitle("ATAC")
p3 <- DimPlot(EC, reduction='wnn.umap') + ggtitle("WNN")
g <- (p1 | p2 | p3) & theme(plot.title = element_text(hjust = 0.5))
print(g)
```

```{r}
FeaturePlot(
  EC,
  features = c("RNA.weight", "ATAC.weight"),
  reduction = "wnn.umap"
)
```

### Cell annotation

```{r}

Idents(EC) <- "cell_type_automatic"

# plot
p1 <- DimPlot(EC, reduction = "umap_harmony", group.by = "cell_type_automatic", label = TRUE, label.size = 2.5) + ggtitle("RNA")
p2 <- DimPlot(EC, reduction = "umap.atac", group.by = "cell_type_automatic", label = TRUE, label.size = 2.5) + ggtitle("ATAC")
p3 <- DimPlot(EC, reduction = "wnn.umap", group.by = "cell_type_automatic", label = TRUE, label.size = 2.5) + ggtitle("WNN")

g <- (p1 | p2 | p3) & NoLegend() & theme(plot.title = element_text(hjust = 0.5))

print(g)

```

## Regolatory - Identificatin of enriched motifs

```{r}

suppressPackageStartupMessages({
  library(chromVAR)
  library(JASPAR2020)
  library(TFBSTools)
  library(motifmatchr)
  library(future)
  library(BSgenome.Hsapiens.UCSC.hg38)
})

```

```{r}

# EC_subset <- subset(
#   EC,
#   subset = cell_type_automatic %in% c("colon epithelial cell", "intestine goblet cell", "macrophage", "glial cell")
# )


DefaultAssay(EC) <- "ATAC"

EC_set <- getMatrixSet(x = JASPAR2020, opts = list(species = 9606, all_versions = FALSE))

motif.matrix <- CreateMotifMatrix(features = granges(EC), pwm = EC_set, genome = 'hg38', use.counts = FALSE)

motif.object <- CreateMotifObject(data = motif.matrix, pwm = EC_set)

EC <- SetAssayData(EC, 
                     assay = 'ATAC', 
                     layer = 'motifs', 
                     new.data = motif.object)

# Note that this step can take 30-60 minutes
#plan("sequential") 
#plan(multisession(workers = 8))


```

```{r}

library(BiocParallel)

register(MulticoreParam(8)) # Use 8 cores

# For Windows, MulticoreParam will not work, but you can use SnowParam:


# register(SerialParam())
```

```{r}

EC_chromvar <- RunChromVAR(
  object = EC,
  genome = BSgenome.Hsapiens.UCSC.hg38
)
```

### Finding TF for each cell type

**CDX1 / CDX2:** Intenstial epithelial linage determination

**TCF4:** secretory/goblet linage

**SPDEF:** goblet cell maturation + mucin production

**XBP1**: goblet cell function - controls muscus barrier in colon by regulating MUC2

**HNF1A:** epithelial differentiaton and metabolims (influence growth ect.)

**STAT:** inflammaotry activation [https://pmc.ncbi.nlm.nih.gov/articles/PMC12151090/](https://pmc.ncbi.nlm.nih.gov/articles/PMC12151090/?utm_source=chatgpt.com)

**YAP-TEAD4** - in colonic macrophages tissue repairCDX1 / CDX2

<https://pubmed.ncbi.nlm.nih.gov/40189621/>

**ELF4:** anti-inflammatory macrophages

Find på glial cells!

```{r}
DefaultAssay(EC_chromvar) <- "chromvar"

motif.name <- ConvertMotifID(EC, name = 'CDX1')
gene_plot <- FeaturePlot(EC, features = "CDX1", reduction = 'wnn.umap')

motif_plot <- FeaturePlot(EC_chromvar, features = motif.name, min.cutoff = 0, cols = c("lightgrey", "darkred"), reduction = 'wnn.umap')

p3 <- DimPlot(EC_chromvar, reduction = "wnn.umap", group.by = "cell_type_automatic", label = TRUE, label.size = 2.5, repel = TRUE) + ggtitle("WNN") + NoLegend()
gene_plot | motif_plot | p3
```
